Monitoring and Projecting Global Hunger: Are We On Track?

### 2021-06-11 docs/response_to_reviewers.md

---
created: 2021-06-10T10:04:08-0400
modified: 2021-06-11T16:36:24-0400
title: Response to Reviewers
header-includes:
 - \hypersetup{colorlinks=true}
 - \hypersetup{urlcolor=blue}
---

@@

Third, Random Forrest models suffer from training data bias. If the underlying training data if flawed, the Random Forest Model has no way of identifying this. What quality control is used to determine if your survey data is correct? One way to address this is to compare the Gallup data with other sources like DHS or LSMS surveys from the same year. Do they both measure similar national-level food security outcomes? I am also curious to know how your models preform with simple linear or logistic regression? Do you get the same results? I recognize Random Forest are nonparametric, but I would be keen to see how linear models preform with controls. Also, what happens when you trade out wasting/stunting as your predicted outcome versus the FIES-based metric?

Fourth, the spatial nature of your modeling is confusing and not clear. Are metrics averaged at the sub-national scale (if so what level of analysis and for which country)? How is degree of urbanization quantified (e.g. are Worldpop pixels masked at some threshold?)? Is the Gallup representative at the national level? District-level? What controls are used to ensure that the Gallup data is a representative sample (Not even explained in the Supplemental Materials)? Your methodology (including how you address issues of spatial scale) needs to be more clearly described in the Main Text and the Supplemental Material before this paper can be considered for publication.

**To add more clarity, we have changed the phrasing in most places from "subnational" to "administrative area", as this might improve some of the confusion around the spatial scale of our analysis.**

@@

https://userforum.dhsprogram.com/index.php?t=msg&th=7832&start=0&

Fifth, I take issue with this being the first global food security paper at a sub-national level (Pg 12, ln 11). Niles et al. is another similar paper, that is near global, with a much more robust modeling framework. Additionally, FEWS-net and WFP provide these assessments regularly or real time. Specifically, you do not present any sub-national findings. What are they? Why is sub-national important and how do your results advance our understanding of food security over space and time? Surely there may be important nuances in the sub-national data (e.g. Does Nairobi look like Dhaka? Is your assumption that two households with similar characteristics, irrespective of geography, correct? Do your results show that this assumption is correct?).

**When stating that this was the first global analysis, we meant that this was the first analysis using the FIES metric, and have ensured that anywhere we make that claim throught the paper, we add that important caveat.  We also believe that our use of the word "global" indicates that our analysis has estimated food insecurity in all countries or nearly all countries, which was not done by Niles et al (who only used 19 countires) and is not done by the FEWSNET reports (which only targets certain developing regions).  Thus, we believe that this claim is valid under our interpretation, and we believe future readers would interpret the word "global" the same way.  However, if you disagree, please let us know in a future review and we can change the language.**

**Our lacking in interpretation of subnational findings was a real oversight, and we have added interpretation around that to the results and discussion sections.**

### 2021-07-08 docs/response_to_reviewers.md

---
created: 2021-06-10T10:04:08-0400
modified: 2021-07-08T14:25:54-0400
title: Response to Reviewers
header-includes:
 - \hypersetup{colorlinks=true}
 - \hypersetup{urlcolor=blue}
---

@@

Second, like the population data employed, I have strong concerns with the temperature and precipitation data and methodology employed here. Why use TerraClimate over the widely used and higher resolution CHIRPS (Funk et al. 2015) or other climate reanalysis (ERA5, MERRA-2, etc.) and the new CHIRTS-Max (Funk et al 2019) or CHIRTS-daily (Verdin et al. 2020) high-resolution temperature products to establish your baseline? What temporal resolution is 'mean temperature'? Surely it is widely established the annualized climate data over large distance data does not reflect the true nature of coupling between household food security outcomes and climate (see Shukla et al. 2020 for one recent example). I point to the example of Sternberg 2012 that clearly shows how climate shocks/variability across the planet can telescope to food security outcomes at great distances. How climate (across spatial and temporal scales) affect local price and labor is key here, yet you do not even discuss price at all, much less employ it, in your modeling paradigm. Yet we know that specifically for urban areas, where half the planet lives, food security outcomes are highly dependent on labor and price (see Blekking et al. 2020 for an example). Also, what about seasonality (again see Shukla et al. 2020 for one example)? None of these crucial aspects of household-level food security outcomes are clearly addressed in the current version of the manuscript. I suspect that the reason your 'climate' variables do not predict your outcome is because you spatially and temporally smooth your data to the point where it will no matter what not capture the true nature of how climate shocks impact food security (e.g. huge spatiotemporal coupling between where food is grown, when it is grown, how much the price is dictated by local, national, and global market forces, etc.). See Niles et al. 2021 for a much more robust approach.

**We thank you for this detailed feedback, and we agree that the approach to modeling the impacts of climate on food security is weak.  However, we are in many ways constrained by our survey data, which is only available at the annual temporal scale and only reported at the national spatial scale, which we use to infer subnational scales.  This makes it much harder to trace the impacts of climate on food security and necessitates the smoothing of our climate data.  The papers you cite by Niles, Blekking, and Shukla all seem to have indicators of food security available at fine spatial scales (Niles uses DHS clusters/community GPS points, Blekking uses household GPS points) and fine temporal scales (Niles uses monthly data, Shukla uses Dekads).**

**We also note that it is important to consider the scope of our analysis.  Weather shocks as well as labor and food prices certainly explain a lot of the variation in food security between households and across seasons.  However, the scope of this analysis is to explore drivers of food insecurity across the globe and over decadal timeframes.  We are especially interested in modeling future trends in food security at these broad spatial and temporal scales.  We therefore use GCM projections, which are useful insofar as they are indicative of shifting baselines, but are unable to anticipate actual droughts or floods that would impact food insecurity in 2030.**

**Given these considerations of how the scale of our data (annual, national scale) and scope of our analysis (modeling food insecurity globally and out to 2030) makes it infeasible to explore the effect of climate on food insecurity beyond shifting baselines, we do not think that using other datasets like ERA-5 or MERRA-2 would add much value. (Also, the CHIRPS and CHIRTS datasets, while excellent, are not truly global and unavailable for certainly latitudes, so couldnt be applied to this global analysis).  However, you raise very important considerations in your feedback and we therefore make two improvements to the paper: (1) in addition to the survey year's rainfall totals and precipitation averages, we add a feature for the previous year's total rainfall and mean temperature, and additionally (2) we add a paragraph to the discussion highlighting many of the points you raise and further outlining the considerations that led to the analytical approach we took.**

@@

Third, Random Forrest models suffer from training data bias. If the underlying training data if flawed, the Random Forest Model has no way of identifying this. What quality control is used to determine if your survey data is correct? One way to address this is to compare the Gallup data with other sources like DHS or LSMS surveys from the same year. Do they both measure similar national-level food security outcomes? I am also curious to know how your models preform with simple linear or logistic regression? Do you get the same results? I recognize Random Forest are nonparametric, but I would be keen to see how linear models preform with controls. Also, what happens when you trade out wasting/stunting as your predicted outcome versus the FIES-based metric?

**Unforuntately, the Gallup World Poll (GWP) is currently the only source of nationally-respresntative data on the FIES, as this metric is not collected by surveys like the DHS and LSMS (although as the FIES has been adopted as an SDG indicator, this is changing, see [here](https://userforum.dhsprogram.com/index.php?t=msg&th=7832&start=0&)).  Thus, we cannot use those datasets to validate our Random Forest model.  However, we note that we do take care to conduct cross validtion at the country level.  We leave out all surveys from 10% of countries when evaluating our model, and find our model it still performs favorably.**

**Additionally, we have updated the paper to now model the FIES using linear regression and lasso regression (in the supplement), and note that, compared to logisitic regression, the random forest model performs quite well.  We also now add some discussion around the importance of accounting for interactions between covariates based on a comparison between random forest and linear regression.**

@@

Fourth, the spatial nature of your modeling is confusing and not clear. Are metrics averaged at the sub-national scale (if so what level of analysis and for which country)? How is degree of urbanization quantified (e.g. are Worldpop pixels masked at some threshold?)? Is the Gallup representative at the national level? District-level? What controls are used to ensure that the Gallup data is a representative sample (Not even explained in the Supplemental Materials)? Your methodology (including how you address issues of spatial scale) needs to be more clearly described in the Main Text and the Supplemental Material before this paper can be considered for publication.

**To add more clarity, we have changed the phrasing in most places from "subnational" to "administrative area", as this might improve some of the confusion around the spatial scale of our analysis.  The Gallup data is a purposive sample designed to be representative at the national level, and we dissaggregated it based on demographic characteristics to the administrative area level.  Gallup records whether individual respondents are in urban or rural areas based on the individuals own reporting.  We then use global gridded data that is available under various SSPs to estimate the percent of the population in urban areas for each administrative area.  We tried to cover this in the _Disaggregation_ subsection of the _Methods_ section, and have endeavoured to make this section even more clear and explicit.**

### 2021-07-13 docs/modeling_fies.tex
finish response to reviewers, add new LASSO analysis

@@

%\maketitle
\begin{abstract}

This paper presents the first global picture of food security at a subnational level based on the Food Insecurity Experience Scale, the indicator for the Sustainable Development Goal of ``Zero Hunger" that is most indicative of the individual's lived experience of food insecurity and hunger.  Using microdata from 75 countries and filling gaps using machine learning, we find significant heterogeneity in levels of food insecurity around the world.  Examining global temporal trends and accounting for the effects of the COVID-19 pandemic, we find that rates of severe food insecurity are declining, resulting in global decreases in the total number of severely food insecure people.  However, the total number of moderately food insecure people has been increasing and, after recovering from the shock of the COVID-19 pandemic, we predict it will continue to increase through the end of the 2020s.  Overall, we conclude that current trends in development and demographic change will still leave a large share of the world's population still experiencing hunger by 2030.  
\end{abstract}

@@

Drawing on the insights derived in designing and implementing these novel food security metrics, the FIES was developed by the Food and Agricultural Organization (FAO) of the United Nations (UN) \cite{Ballard2013} and is now recognized as a rapidly deployable and cross-culturally valid tool for understanding individual and household-level food insecurity \citep{wambogo2018validity, smith2017world}.  The FIES is based on a survey of eight behaviors indicative of food insecurity and hunger over the previous year, such as skipping meals or worrying about having enough to eat.  Processing the responses given to the FIES questions through the Rasch model, each individual in the survey is assigned a score that allows them to be ranked on a scale of food insecurity severity. Once the severity measures are calibrated against a common reference (the global FIES reference scale) \citep{Cafiero2018} and individuals are ranked in this fashion, conventional thresholds for ``moderate'' and ``severe'' food insecurity are set and the prevalence of such food insecurity conditions can be computed. Thresholds are set in such a way that a moderately food insecure respondent is one who likely has compromised on food quality and variety as a result of lack of resources, has been unsure about their ability to obtain food and has skipped meals or run out of food occasionally. A severely food insecure one is likely to have run out of food and has gone an entire day without eating at times during the year.  The percentage of people estimated to be over the threshold set for moderate food insecurity includes people that are also over the threshold for severe food insecurity and represents the SDG indicator 2.1.2.

@@

\section{Methods}
\subsection{Disaggregation}
The data on the FIES collected by Gallup on behalf of FAO also records many individual-level attributes from the respondents, including age, gender, wealth quintile, and whether they live in an urban or rural area.  Using national data on these variables, a standard weighting scheme is created for each individual, based on the ratio of population probabilities to sample probabilities of an individual with those characteristics being selected \citep{bethlehem2009applied}.  We use a similar methodology and calculate post-stratification weights at a subnational level.  We use year-specific subnational estimates of population shares by gender, age, urbanization, and, where available, wealth, and create a separate set of weights for each subnational area across all individuals.  For age and gender, we used gridded data from WorldPop \citep{Tatem2017}; for urbanization, we used spatially explicit estimates published by Jones and O'Neill \citep{Jones2016}; and for wealth we used data on household wealth quintiles from Demographic and Health Surveys \citep{dhsall}.  This methodology rests on the mild assumption that, within a given country, two individuals of the same age, gender, urbanization level and wealth quintile will have a similar food security status.

@@

For the climatic variables, temperature and precipitation, we combine historical observations with an ensemble of four bias-corrected simulations of the future climate from the Inter-Sectoral Impact Model Intercomparison Project (ISIMIP) \citep{warszawski2014inter}.

Finally, we adjust several variables to account for the effects of the COVID-19 pandemic.  These include June 2020 GDP predictions from the World Bank \citep{prospects2020}, August 2020 estimates of impacts on anthropometry from \textit{The Lancet} \citep{headey2020impacts}, as well as the Poverty Headcount Index \cite{Cuaresma2018}.

\begin{table}[H]
  \centering
	\begin{tabular}{lll}
		\toprule
    Name & Source & Scale \\
		\midrule
    Urban Percentage & \cite{Jones2016} & Subnational \\ [1cm]
    Stunting & \cite{Local2020} & Subnational \\ [1cm]
    Wasting & \cite{Local2020} & Subnational \\ [1cm]
    Mean Years of Schooling & \makecell[tl]{\cite{Smits2019} \\ \cite{KC2017}} & Subnational\\ [1cm]
    GDP Per Capita & \makecell[tl]{\cite{Smits2019} \\ \cite{Dellink2017}} & Subnational \\ [1cm]
    Gini Coefficient & \cite{Rao2019a} & National\\ [1cm]
    Poverty Headcount Index & \cite{Cuaresma2018} & National \\ [1cm]
    Water Scarcity & \cite{greve2018global} & Subnational \\ [1cm]
    Mean Annual Precipitation & \makecell[tl]{\cite{abatzoglou2018terraclimate} \\ \cite{warszawski2014inter}} & Subnational \\ [1cm]
    Topographic Ruggedness &  \makecell[tl]{\cite{USGS1996} \\ \cite{Riley1999}} & Subnational \\ [1cm]
    Mean Temperature &  \makecell[tl]{\cite{abatzoglou2018terraclimate} \\ \cite{warszawski2014inter}} & Subnational \\ [1cm]
    \makecell[tl]{Malaria (\textit{P. falciparum}) \\ Mortality Rate} & \cite{Weiss2019} & Subnational \\
		\bottomrule
	\end{tabular}
	\caption{Covariates Included in the Analysis}
	\label{tab:covars}
\end{table}

\subsection{Modeling}
We fit two models for all subnational areas globally from 2010 to 2030: one for the percent of the population over the threshold for moderate food insecurity, and one for the percent of the population over the threshold for severe food insecurity.  We use random forest regressions, which perform well in the presence of non-linearities and interaction effects among predictor variables \citep{hastie2009elements}.  A random forest regression involves generating a large number of decision trees, each using different sub-samples of the data, and then aggregating the predictions of the decision trees.  By taking an ensemble of decision tree models, random forests introduce more variance and balance out the bias that is common to methods based on single decision trees \citep{friedman2001elements}. Given the bounded nature of our variables of interest, the random forest regression are applied to the logistically transformed rates.  We create our models using the \texttt{randomForestSRC} package in R \citep{ishwaran2019randomforestsrc}.

To get the best possible model performance, it is necessary to tune the hyperparamters, which control the process in which the random forest algorithm is run.  These include the average number of observations in a leaf node, the number of variables randomly selected as candidates for splitting a node, and the maximum depth of each tree.  We use the combination of hyperparameters that perform best under leave-one-out validation based on $R^2$.

When performing cross validation, it is typical to select observations at random.  However, in our training data we have multiple highly similar observations from each country across years and subnational areas, and we extend our model predictions to countries that were not found at all in our training data.  Thus, we perform a type of leave-one-out cross validation, where we leave out all of the data from one country for each iteration.  Thus, the model error after cross validation gives us a sense of how the model is performing in countries where FIES data was not collected.  By this validation metric, we find that our models perform well, with an $R^2$ of 0.792 and mean average error (MAE) of 0.071 for moderate food insecurity and an $R^2$ of 0.746 and a MAE of 0.041 for severe food insecurity.  After fitting models on the entire dataset, we observe a within-sample $R^2$ of 0.950 and an MAE of 0.034 for moderate food insecurity and an $R^2$ of 0.945 and a MAE of 0.018 for severe food insecurity.

@@

\section{Results}
\subsection{Food Insecurity over Time}
Globally, in the year 2020, we estimate that 820 million people are above the threshold for severe food insecurity, and 2.5 billion people are above the threshold for moderate food insecurity (See Fig. \ref{fig:timeseries}). These numbers are higher than the estimates in the 2020 FAO State of Food Insecurity (SOFI) report, which reports about 746 million severely food insecure people and 2 billion people experiencing at least moderate food insecurity in 2019 \citep{sofi2020}.

Under realistic scenario assumptions, we find divergent trends in the number of moderate and severely food insecure people around the world, with the number of people at least moderately food insecure increasing and the number of people severely food insecure declining.  However, for both indicators, these changes are very slight: in 2030, the number of moderately food insecure people will have increased by 2.2\% since 2010, and the number of severely food insecure people will have decreased by 7.0\%.  The increases in moderate food insecurity have been in lower and lower-middle income countries, while the decreases in severe food insecurity are entirely from lower income countries.  For both the moderate and severe food insecurity thresholds, COVID-19 has had a clearly discernible impact, particularly in middle-income countries.

@@

\subsection{Food Insecurity Across Space}
We find substantial heterogeneity in the global distribution of severe food insecurity across countries and subnational regions (See Fig. \ref{fig:map}).  For the year 2020, mainland sub-Saharan Africa is the continent with the highest rates of severe food insecurity, with 20\% of people over the corresponding threshold in at least one subnational area in every country except Gabon, Djibouti and Equatorial Guinea.  Outside of Africa, serious pockets of severe food insecurity also occur in Venezuela, Syria, Papua New Guinea, Yemen, and Afghanistan.  In many large middle-income countries, severe food insecurity is also quite prevalent, with rates between 10\% and 15\% in northern Brazil, many central Asian and middle-eastern countries, as well as India and Indonesia.

@@

@@

\subsection{Important Predictors of Food Insecurity}
Overall, our model was able to predict food insecurity with high accuracy in the countries for which training data was available ($R^2 = 0.950$ and $R^2 = 0.940$ for moderate and severe food insecurity, respectively), and excluding entire countries from the training data to evaluate out-of-country model error still showed overall good results ($R^2 = 0.792$ and $R^2 = 0.746$ for moderate and severe food insecurity, respectively).  The relative importance of different variables as predictors of food insecurity can be assessed by examining the loss in forecast accuracy resulting from the exclusion of particular covariates.

For severe food insecurity, the most important variable for predicting food security levels is the Poverty Headcount Index (See Fig. \ref{fig:vimp}), followed by the rate of stunting and the Gini coefficient.  For predicting the rate of people with at least moderate food insecurity, GDP per capita is the most important, followed closely by the Poverty Headcount Index.  The country's Gini coefficient is also important, more so for predicting the rate of severe food insecurity compared to moderate food insecurity.  Across both outcomes, economic variables were the most relevant predictors.

@@

Our results anticipating increases in the number of people over the threshold for moderate food insecurity concur with the FAO's predictions of increased rates of undernourishment, the other indicator for the SDG 2 target of ending hunger \citep[Part 1, Page 11]{sofi2020}.  However, severe food insecurity may be a more apt indicator to compare with undernourishment, as the global totals are more similar \citep{sofi2020}.  In this case, our predictions diverge: the FAO estimates a 24\% increase, from 687 to 841 million undernourished people, from 2019 to 2030, whereas we predict a 2.6\% decrease, from 787 to 767 million severely food insecure people over the same period.

These discrepancies are in part due to our differing modeling approaches. While the SOFI report extrapolated recent trends in undernourishment and stunting, we modeled food insecurity based on interactions between projections in factors like GDP per capita, population, and the prevalence of stunting. Additionally, our diverging predictions for different hunger indicators are not necessarily mutually exclusive: it is theoretically possible for undernourishment to increase while severe food insecurity, as measured by the FIES, decreases.  The discrepancies between our projections and those of the FAO illustrate both the complexity of hunger as a social phenomenon, as well as the challenges in predicting future socioeconomic developments, where different modeling approaches may yield different forecasts.

@@

While we find that rates of food insecurity are expected to decrease overall by 2030, our model, which relies on middle-of-the-road assumptions, still expects the number of people experiencing moderate food insecurity to increase and the number of people experiencing severe food insecurity to fall only slightly between 2010 and 2030.  This still leaves billions of people eating less than they should and three-quarters of a billion people going entire days without eating a decade from now, falling far short of the SDG 2 goal of ending hunger.  Moreover, decreases in food insecurity as measured by the FIES do not necessarily mean there will be improvements in other significant challenges such as poor dietary quality, micronutrient deficiencies, sustainable diets, or obesity.  Thus, while expected trends in correlates of food security give cause for optimism, there is still significant work to be done at the political level in pursuit of SDG 2.

\section{Acknowledgements}
This reserach was supported by The International Fund for Agricultural Development (IFAD) grant number 2000002044.  We thank Kristofer Hamel and Wolfgang Fengler for their support in this work.

>>>>

% Use datatool to read in reported values
% csv must have columns `var` and `val`
\usepackage{datatool}
\DTLsetseparator{,}
\DTLloaddb{data}{vars.csv}
\newcommand{\myvar}[1]{\DTLfetch{data}{var}{#1}{val}}

@@

%\maketitle
\begin{abstract}
This paper presents the first global picture of food security at a subnational level based on the Food Insecurity Experience Scale, the indicator for the Sustainable Development Goal of ``Zero Hunger" that is most indicative of the individual's lived experience of food insecurity and hunger.  Using microdata from 75 countries and filling gaps using machine learning, we find significant heterogeneity in levels of food insecurity around the world.  Examining global temporal trends and accounting for the effects of the COVID-19 pandemic, we find that rates of severe food insecurity are declining, resulting in global decreases in the total number of severely food insecure people.  However, the total number of moderately food insecure people has been increasing and, after recovering from the shock of the COVID-19 pandemic, we predict it will continue to increase through the end of the 2020s.  Overall, we conclude that current trends in development and demographic change will still leave a large share of the world's population still experiencing hunger by 2030.  \end{abstract}

@@

Drawing on the insights derived in designing and implementing these novel food security metrics, the FIES was developed by the Food and Agricultural Organization (FAO) of the United Nations (UN) \citep{Ballard2013} and is now recognized as a rapidly deployable and cross-culturally valid tool for understanding individual and household-level food insecurity \citep{wambogo2018validity, smith2017world}.  The FIES is based on a survey of eight behaviors indicative of food insecurity and hunger over the previous year, such as skipping meals or worrying about having enough to eat.  Processing the responses given to the FIES questions through the Rasch model, each individual in the survey is assigned a score that allows them to be ranked on a scale of food insecurity severity. Once the severity measures are calibrated against a common reference (the global FIES reference scale) \citep{Cafiero2018} and individuals are ranked in this fashion, conventional thresholds for ``moderate'' and ``severe'' food insecurity are set and the prevalence of such food insecurity conditions can be computed. Thresholds are set in such a way that a moderately food insecure respondent is one who likely has compromised on food quality and variety as a result of lack of resources, has been unsure about their ability to obtain food and has skipped meals or run out of food occasionally. A severely food insecure one is likely to have run out of food and has gone an entire day without eating at times during the year.  The percentage of people estimated to be over the threshold set for moderate food insecurity includes people that are also over the threshold for severe food insecurity and represents the SDG indicator 2.1.2.

@@

\section{Methods}
This analysis had three steps.  First, in countries where FIES microdata was available at the national level, we generate subnational weights to disaggregate the microdata and estimate food insecurity rates at the first administrative level for each survey year.  Second, we gather data on covariates related to food insecurity from previously published datasets that include future predictions under the Shared Socio-Economic Pathways framework.  These covariates are processed to be at the same yearly, administrative area level, for all years from 2010 to 2030.  Finally, we fit a machine learning model using the covariates in administrative areas where microdata is available, and use that model to make predictions for all years and all administrative areas.

\subsection{Disaggregation}
The data on the FIES collected by Gallup on behalf of FAO also records many individual-level attributes from the respondents, including age, gender, wealth quintile, and whether they live in an urban or rural area.  Using national data on these variables, a standard weighting scheme is created for each individual, based on the ratio of population probabilities to sample probabilities of an individual with those characteristics being selected \citep{bethlehem2009applied}.  This ensures that the results are representative at a national level.  We use a similar methodology, except that we calculate post-stratification weights at a subnational level.  We use year-specific subnational estimates of population shares by gender, age, urbanization, and, where available, wealth, and create a separate set of weights for each subnational area across all individuals.  For age and gender, we used gridded data from WorldPop \citep{Tatem2017}; for urbanization, we used spatially explicit estimates of population in urban and rural pixels published by Jones and O'Neill and deriving from night-time lights and buffered settlement centroids \citep{Jones2016}; and for wealth we used data on household wealth quintiles from Demographic and Health Surveys \citep{dhsall}.  This methodology rests on the mild assumption that, within a given country, two individuals of the same age, gender, urbanization level and wealth quintile will have a similar food security status, an assumption that underlies much of survey weighting methodology \citep{bethlehem2009applied}.

@@

For the climatic variables, mean annual temperature and total annual precipitation, we combine historical observations with an ensemble of four bias-corrected simulations of the future climate from the Inter-Sectoral Impact Model Intercomparison Project (ISIMIP) \citep{warszawski2014inter}.  These variables cannot speak to sudden shocks but do capture the climatic baselines relevant to agro-ecological conditions that constrain food production.

Finally, we adjust several variables to account for the effects of the COVID-19 pandemic.  These include June 2020 GDP predictions from the World Bank \citep{prospects2020}, August 2020 estimates of impacts on anthropometry from \textit{The Lancet} \citep{headey2020impacts}, as well as the Poverty Headcount Index \citep{Cuaresma2018}.

\begin{table}[H]
  \centering
	\begin{tabular}{lll}
		\toprule
    Name & Source & Scale \\
		\midrule
    Urban Percentage & \cite{Jones2016} & Subnational \\ [1cm]
    Stunting & \cite{Local2020} & Subnational \\ [1cm]
    Wasting & \cite{Local2020} & Subnational \\ [1cm]
    Mean Years of Schooling & \makecell[tl]{\cite{Smits2019} \\ \cite{KC2017}} & Subnational\\ [1cm]
    GDP Per Capita & \makecell[tl]{\cite{Smits2019} \\ \cite{Dellink2017}} & Subnational \\ [1cm]
    Gini Coefficient & \cite{Rao2019a} & National\\ [1cm]
    Poverty Headcount Index & \cite{Cuaresma2018} & National \\ [1cm]
    Water Scarcity & \cite{greve2018global} & Subnational \\ [1cm]
    \makecell[tl]{Total Annual Precipitation \\ (Current and Previous Year)} & \makecell[tl]{\cite{abatzoglou2018terraclimate} \\ \cite{warszawski2014inter}} & Subnational \\ [1cm]
    Topographic Ruggedness &  \makecell[tl]{\cite{USGS1996} \\ \cite{Riley1999}} & Subnational \\ [1cm]
    \makecell[tl]{Mean Temperature \\ (Current and Previous Year)} &  \makecell[tl]{\cite{abatzoglou2018terraclimate} \\ \cite{warszawski2014inter}} & Subnational \\ [1cm]
    \makecell[tl]{Malaria (\textit{P. falciparum}) \\ Mortality Rate} & \cite{Weiss2019} & Subnational \\
		\bottomrule
	\end{tabular}
	\caption{Covariates Included in the Analysis}
	\label{tab:covars}
\end{table}

\subsection{Modeling}
We fit two models for all subnational areas globally from 2010 to 2030: one for the percent of the population over the threshold for moderate food insecurity, and one for the percent of the population over the threshold for severe food insecurity.  We use both random forest regressions and LASSO (Least Absolute Shrinkage and Selection Operator) regressions, with the results of the LASSO regressions presented in the supplement.   A random forest regression involves generating a large number of decision trees, each using different sub-samples of the data, and then aggregating the predictions of the decision trees.  This approach performs well in the presence of non-linearities and interaction effects among predictor variables \citep{hastie2009elements}.  By taking an ensemble of decision tree models, random forests introduce more variance and balance out the bias that is common to methods based on single decision trees \citep{friedman2001elements}. Given the bounded nature of our variables of interest, the random forest regression are applied to the logistically transformed rates.  We create our models using the \texttt{randomForestSRC} package in R \citep{ishwaran2019randomforestsrc}.

To get the best possible model performance, it is necessary to tune the hyperparamters, which control the process in which the random forest algorithm is run.  These include the average number of observations in a leaf node, the number of variables randomly selected as candidates for splitting a node, and the maximum depth of each tree.  We use the combination of hyperparameters that perform best under leave-one-out validation based on $R^2$ and mean average error (MAE).

When performing cross validation, it is typical to select observations at random.  However, in our training data we have multiple highly similar observations from each country across years and subnational areas, and we extend our model predictions to countries that were not found at all in our training data.  Thus, we perform a type of leave-one-out cross validation, where we leave out all of the data from one country for each iteration.  Thus, the model error after cross validation gives us a sense of how the model is performing in countries where FIES data was not collected.  By this validation metric, we find that our models perform well, with an $R^2$ of \myvar{rf_cv_mod_rsq} and mean average error (MAE) of \myvar{rf_cv_mod_mae} for moderate food insecurity and an $R^2$ of \myvar{rf_cv_sev_rsq} and a MAE of \myvar{rf_cv_sev_mae} for severe food insecurity.  After fitting models on the entire dataset, we observe a within-sample $R^2$ of \myvar{rf_wthn_mod_rsq} and an MAE of \myvar{rf_wthn_mod_mae} for moderate food insecurity and an $R^2$ of \myvar{rf_wthn_sev_rsq} and a MAE of \myvar{rf_wthn_sev_mae} for severe food insecurity.

@@

\section{Results}
\subsection{Food Insecurity over Time}
Globally, in the year 2020, we estimate that \myvar{glob_2020_sev} million people are above the threshold for severe food insecurity, and \myvar{glob_2020_mod} billion people are above the threshold for moderate food insecurity (See Fig. \ref{fig:timeseries}). These numbers are higher than the estimates in the 2020 FAO State of Food Insecurity (SOFI) report, which reports about 746 million severely food insecure people and 2 billion people experiencing at least moderate food insecurity in 2019 \citep{sofi2020}.

Under realistic scenario assumptions, we find recent increases in the number of moderate and severely food insecure people around the world, yet expect overall decreases from 2010 to 2030.  These changes are more slight for people who are at least moderately food insecure: in 2030, the number of moderately food insecure people will have decreased by \myvar{decprc_2010_2030_mod}\% since 2010, and the number of severely food insecure people will have decreased by \myvar{decprc_2010_2030_sev}\%.  While the number of moderately and severely food insecure people are modeled to decrease overall, they are projected to increase in lower income countries through 2030.

@@

\subsection{Food Insecurity Across Space}
We find substantial heterogeneity in the global distribution of severe food insecurity across countries and subnational regions (See Fig. \ref{fig:map}).  For the year 2020, mainland sub-Saharan Africa is the continent with the highest rates of severe food insecurity, with 20\% of people over the corresponding threshold in at least one subnational area in every country except Gabon, Djibouti and Equatorial Guinea.  Outside of Africa, serious pockets of severe food insecurity also occur in Venezuela, Syria, Papua New Guinea, Yemen, and Afghanistan.  In many large middle-income countries, severe food insecurity is also quite prevalent, with rates between 10\% and 15\% in parts of Latin America, many central Asian and middle-eastern countries, as well as India and Indonesia.

@@

Some countries show persistent subnational gaps in rates of food insecurity, typically in areas that are more remote.  For example, the Somali province of Ethiopia, Sahelo-Saharan Mali, and Amazonian Brazil are all generally more food insecure subnational areas than the more populous and accessible parts of their respective countries.

@@

\subsection{Important Predictors of Food Insecurity}
Overall, our model was able to predict food insecurity with high accuracy in the countries for which training data was available ($R^2 = \myvar{rf_wthn_mod_rsq}$ and $R^2 = \myvar{rf_wthn_sev_rsq}$ for moderate and severe food insecurity, respectively), and excluding entire countries from the training data to evaluate out-of-country model error still showed overall good results ($R^2 = \myvar{rf_cv_mod_rsq}$ and $R^2 = \myvar{rf_cv_sev_rsq}$ for moderate and severe food insecurity, respectively).  The relative importance of different variables as predictors of food insecurity can be assessed by examining the loss in forecast accuracy resulting from the exclusion of particular covariates.

For severe food insecurity, the most important variable for predicting food security levels is the Poverty Headcount Index (See Fig. \ref{fig:vimp}), followed by the Gini coefficient and the rate of stunting.  For predicting the rate of people with at least moderate food insecurity, GDP per capita is the most important, followed closely by the Poverty Headcount Index.  The country's Gini coefficient is also important, more so for predicting the rate of severe food insecurity compared to moderate food insecurity.  Across both outcomes, economic variables were the most relevant predictors.

@@

It should be noted that our model was using covariates and making predictions a coarser scale - the administrative area - than much of the food security literature, which, since the 1990s, has increasing focused at the scale of the household \citep{Cooper2020}.  Thus, this analysis cannot speak to the localized impact of factors like climate, urbanization, or topography at the finer spatio-temporal scales at which these variables may operate \citep{Shukla2021, Blekking2020, Cockx2018, Tuholske2020}.  For example, our predictor variables for temperature and precipitation, derived from future simulations of the climate, can only model the effects of shifting baselines as data is smoothed over the course of the year - our model cannot anticipate the effects of short-term and extreme climate events on food insecurity.  Additionally, our model can only capture coarse, administrative area metrics of urbanization and cannot address household-level differences in food security between urban and rural households.  Nevertheless, the fact that these variables explained some of the variation in food security outcomes, even at administrative area scale, underscores their importance for food security.

Our results anticipating decreases in the number of people over the threshold for moderate food insecurity conflict with the FAO's predictions of increased rates of undernourishment, the other indicator for the SDG 2 target of ending hunger \citep[Part 1, Page 11]{sofi2020}.  However, severe food insecurity may be a more apt indicator to compare with undernourishment, as the global totals are more similar \citep{sofi2020}.  In this case, our predictions still diverge: the FAO estimates a 24\% increase, from 687 to 841 million undernourished people, from 2019 to 2030, whereas we predict a \myvar{decrprc_2019_2030_sev}\% decrease, from \myvar{sev_tot_2019} to \myvar{sev_tot_2030} million severely food insecure people over the same period.

These discrepancies are in part due to our differing modeling approaches. Food insecurity was worsening slightly in the latter half of the 2010s, a trend noted elsewhere \cite{Voosen2020}, and reflected in our own analysis (Fig. \ref{fig:timeseries}), and the SOFI report extrapolated these recent trends in undernourishment and stunting.  Our approach, on the other hand, modeled food insecurity based on interactions between projections in underlying drivers like GDP per capita and population change.  Additionally, our diverging predictions for different hunger indicators are not necessarily mutually exclusive, as we are using different indicators: it is theoretically possible for undernourishment to increase while severe food insecurity, as measured by the FIES, decreases.  The discrepancies between our projections and those of the FAO illustrate both the complexity of hunger as a social phenomenon, as well as the challenges in predicting future socioeconomic developments, where different modeling approaches may yield different forecasts.

@@

While we find that rates of food insecurity are expected to decrease overall by 2030, our model, which relies on middle-of-the-road assumptions, still expects the number of people experiencing moderate or severe food insecurity to increase in low income countries and the overall number of people experiencing moderate or severe food insecurity to fall only slightly between 2010 and 2030.  This still leaves billions of people eating less than they should and three-quarters of a billion people going entire days without eating a decade from now, falling far short of the SDG 2 goal of ending hunger.  Moreover, decreases in food insecurity as measured by the FIES do not necessarily mean there will be improvements in other significant challenges such as poor dietary quality, micronutrient deficiencies, sustainable diets, or obesity.  Thus, while expected trends in correlates of food security give cause for optimism, there is still significant work to be done at the political level in pursuit of SDG 2.

\section{Acknowledgements}
This research was supported by The International Fund for Agricultural Development (IFAD) grant number 2000002044.  We thank Kristofer Hamel and Wolfgang Fengler for their support in this work.
